---
title: "Annotation"
author: "Nadine Bestard"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Set-up
```{r library, message=FALSE, warning=FALSE}
library(here) #reproducible paths
library(scran) #aggregate
library(scater) # plot sc
library(readxl) # read the anotation file
```
```{r}
project<- "fire-mice"
source(here("src/colours.R"))
```


## Loading
```{r load}
sce <- readRDS(here("processed", project, "sce_clusters_01.RDS"))
```

## Top genes

For each cluster the top experssed genes are calculated and saved in `r paste("outs/", project, "/top_genes_res0.7.tsv")`
```{r top}

if(!file.exists(here("outs", project, "top_genes_res0.7.tsv"))){
# Create object where instead of cells in cols we have the clusters with counts aggregated
summed_res0.7 <- aggregateAcrossCells(sce, id=colData(sce)[,c("originalexp_snn_res.0.7")])

# # Delete ribosomal and mitochondrial genes
# is_ribo <- grepl("^Rp[sl]", rownames(summed_res0.4))
# is_mito <- grepl("^mt-", rownames(summed_res0.4))
# keep <- !(is_ribo | is_mito)
#  summed_res0.4 <- summed_res0.4[keep,]
 
# for each column (aggregated counts) of the object extract the top genes
top_genes_res0.7 <- apply(counts(summed_res0.7), 
      MARGIN = 2,
      function(x){
        # get the gene names from the top 50 expressed genes
        names(sort(x, decreasing = T)[1:100])
        }
      )
# save result
write.table(top_genes_res0.7, file = here("outs", project, "top_genes_res0.7.tsv"), sep = "\t", quote = FALSE, col.names = TRUE, row.names = FALSE)
}

```

## re-swap samples

We noticed with the preliminary shiny app that one of the samples had been wrongly labelled. 
We have found from where the error came from, two mice were swaped in one the first preliminary stages during the 10x lab protocol.
We correct this mistake here swapping back mouse1 and mouse2. 

This affects the following metadata:

- original_sample_name
- Sample
- mouse
- genotype


```{r swap}
# store the position of the values I want to replace, they are always in the same position, as each row is a cell. 
KOvo1_1 <- sce$original_sample_name == "KOvo1.1"
KOvo1_2 <- sce$original_sample_name == "KOvo1.2"
WTvo1_1 <- sce$original_sample_name == "WTvo1.1"
WTvo1_2 <- sce$original_sample_name == "WTvo1.2"

# replace the wrong values by the new values
# original sample name
sce$original_sample_name <- replace(sce$original_sample_name, KOvo1_1, "WTvo1.1")
sce$original_sample_name <- replace(sce$original_sample_name, KOvo1_2, "WTvo1.2")
sce$original_sample_name <- replace(sce$original_sample_name, WTvo1_1, "KOvo1.1")
sce$original_sample_name <- replace(sce$original_sample_name, WTvo1_2, "KOvo1.2")
# sample
sce$Sample <- replace(sce$Sample, KOvo1_1, "WTvo1_1")
sce$Sample <- replace(sce$Sample, KOvo1_2, "WTvo1_2")
sce$Sample <- replace(sce$Sample, WTvo1_1, "KOvo1_1")
sce$Sample <- replace(sce$Sample, WTvo1_2, "KOvo1_2")
# mouse
sce$mouse <- replace(sce$mouse, KOvo1_1, "very_old_mouse_1")
sce$mouse <- replace(sce$mouse, KOvo1_2, "very_old_mouse_1")
sce$mouse <- replace(sce$mouse, WTvo1_1, "very_old_mouse_2")
sce$mouse <- replace(sce$mouse, WTvo1_2, "very_old_mouse_2")
# genotype
sce$genotype <- replace(sce$genotype, KOvo1_1, "WT")
sce$genotype <- replace(sce$genotype, KOvo1_2, "WT")
sce$genotype <- replace(sce$genotype, WTvo1_1, "KO")
sce$genotype <- replace(sce$genotype, WTvo1_2, "KO")
```

<!-- ``` -->
<!-- #before swap -->
<!-- head(which(sce$original_sample_name %in% c("KOvo1.1"))) -->
<!-- [1] 1 2 3 4 5 6 -->
<!-- head(which(sce$original_sample_name %in% c("KOvo1.2"))) -->
<!-- [1] 1606 1607 1608 1609 1610 1611 -->
<!-- head(which(sce$original_sample_name %in% c("WTvo1.1"))) -->
<!-- [1] 13565 13566 13567 13568 13569 13570 -->
<!-- head(which(sce$original_sample_name %in% c("WTvo1.2"))) -->
<!-- [1] 15873 15874 15875 15876 15877 15878 -->
<!-- ``` -->

# Annotate

With the shiny App, canonical markers were used to annotate the dataset, together with the top 100 genes 
a provisional annotation is given to each cluster.

```{r annotation}
annotation <- read_xlsx(here("data/Very old FIRE mice cluster names resolution 0.7.xlsx"))

sce$celltype <- sce$originalexp_snn_res.0.7
sce$clusters_named <- sce$originalexp_snn_res.0.7

levels(sce$celltype) <- annotation$CellType
levels(sce$clusters_named) <- annotation$ClusterName

sce$genotype <- relevel(factor(sce$genotype), "WT")

plotTSNE(sce, colour_by = "clusters_named", text_by = "clusters_named", text_size = 3, force = 1) + scale_color_manual(values = cols)

```

```{r save}
if (!file.exists(here("processed", project, "sce_anno_01.RDS"))) {
  saveRDS(sce, here("processed", project, "sce_anno_01.RDS"))
}
```

