---
title: "Second QC"
author: "Nadine Bestard"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 80
---

# Set-up

```{r output-code, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r set-up, message=FALSE, warning=FALSE, eval=TRUE }
library(here) # for reproducible paths
library(SingleCellExperiment)
library(scater) # For qcs
library(ggplot2) # To add titles to plots
library(patchwork) # agregate plots
```

```{r load-sce, eval=TRUE}

project <- "fire-mice"
sce <- readRDS(here("processed", project, "sce_clusterQC.RDS"))
source(here("src/colours.R"))
```

The object has `r dim(sce)[1]` genes and `r dim(sce)[2]` cells.

# Cell QC

Previous thresholds were:

|                |               |                 |              |                |          |       |
|----------------|---------------|-----------------|--------------|------------|------------|------------|
|                | lib_size_high | expression_high | lib_size_low | expression_low | mt_pct   | total |
| Cells filtered | 1561          | 96              | 0            | 116            | 5618     | 7291  |
| Threshold      | 17346.98      | 6293.816        | 156.1676     | 108.9684       | 31.10665 | NA    |

<details>

<summary>Click to expand plots</summary>

### Violin plots

Split by sample

```{r violinsample, eval=TRUE}
plotColData(sce, x = "Sample", y = "subsets_mt_percent") +
  scale_y_log10() + ggtitle("mt genes") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  scale_color_manual(values = cols)
plotColData(sce, x = "Sample", y = "detected") +
  scale_y_log10() + ggtitle("Detected genes log scale") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  scale_color_manual(values = cols)
plotColData(sce, x = "Sample", y = "sum") +
  scale_y_log10() + ggtitle("Total count log scale") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  scale_color_manual(values = cols)
```

Split by sample and cluster, facet wrap

```{r violinwrap, fig.height=14, fig.width=14, eval=TRUE}

# detected
plotColData(sce, x = "Sample", y = "detected",  colour_by = "genotype", other_fields = "clusters_named") +
  scale_y_log10(breaks = scales::pretty_breaks(n = 12)) + 
  ggtitle("Detected genes log scale") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +           scale_color_manual(values = col_wt_ko)  + facet_wrap(~clusters_named)
# sum
plotColData(sce, x = "Sample", y = "sum",  colour_by = "genotype", other_fields = "clusters_named") +
 scale_y_log10(breaks = scales::pretty_breaks(n = 12))+
  ggtitle("Total count log scale") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  scale_color_manual(values = col_wt_ko)+ facet_wrap(~clusters_named)
#mt
plotColData(sce, x = "Sample", y = "subsets_mt_percent",  colour_by = "genotype", other_fields = "clusters_named") +
 scale_y_log10(breaks = scales::pretty_breaks(n = 12))+
  ggtitle("Mt. percentage") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  scale_color_manual(values = col_wt_ko)+ facet_wrap(~clusters_named)

```

```{r hist}
hist(
  sce$detected,
  breaks = 100
)
hist(
  sce$sum,
  breaks = 100
)
hist(
  sce$subsets_mt_percent,
  breaks = 100
)
```

### Scatter plots

```{r scatter, fig.height=14, fig.width=14, eval=FALSE}

plotColData(sce, x = "sum", y = "subsets_mt_percent", colour_by = "genotype", other_fields = "clusters_named") + facet_wrap(~clusters_named) + 
  scale_color_manual(values = col_wt_ko)

plotColData(sce, x = "sum", y = "detected", colour_by = "genotype", other_fields = "clusters_named") + facet_wrap(~clusters_named) + 
  scale_color_manual(values = col_wt_ko)

```

```{r scatter2, fig.height=14, fig.width=14, eval=TRUE}

plotColData(sce, x = "sum", y = "subsets_mt_percent", colour_by = "Sample", other_fields = "clusters_named") + facet_wrap(~clusters_named) + 
  scale_color_manual(values = cols)

plotColData(sce, x = "sum", y = "detected", colour_by = "Sample", other_fields = "clusters_named") + facet_wrap(~clusters_named) + 
  scale_color_manual(values = cols)

```



</details>

Sample WT3_2 and WT2_1 are lower quality. When deciding the thresholds they will be ignored. 

We set the thresholds for each celltype:

**Sum**

OPCs 6000 -na

oligo 6000- exept olito 3 - 

Astrocyte 4000 

Microglia and immune 3000

vascular 3000

Ependymocytes&NRPs 2000

mature_neurons 2000

Gran&Mono 2000

ChP_epithelial 4000

**Mt**

all 15% 

```{r subset, eval=TRUE}

# create vector with same levels as the clusters_named to replace by thresholds for each
sce$clusters_named <- droplevels(sce$clusters_named)
sce$threshold_umi <- (sce$clusters_named)
#to get the order:
#dput(levels(sce$threshold_umi))
#c("Astro1", "Astro2", "Oligo1", "Astro3", "Vascular", "OPC", "Microglia1", "Immune", "Oligo2", "mNeurons", "Ependymocytes&NRPs", "Gran&Mono", "Oligo3", "ChP epithelia")

# replace 
levels(sce$threshold_umi) <- c(4000, 4000, 6000, 4000, 3000, 6000, 
3000, 3000, 6000, 2000, 2000, 
2000, 0, 4000)


# transform the factor into numeric
sce$threshold_umi <- as.numeric(as.character(sce$threshold_umi))

# set thresholds
discard_umi <- sce$sum < sce$threshold_umi
discard_mt <- sce$subsets_mt_percent > 15
discard <- discard_umi | discard_mt

# subset
sce <- sce[,!discard]
```

We discard `r sum(discard_umi)` cells due to umi counts,
and `r sum(discard_mt)` cells
for high mt percentatge.

in total we discard `r sum(discard)` cells

# Gene QC

In the preliminary analysis we only deleted genes that had less than 2 cells
expressing it, now we filter all genes with less than 10 cells expressing it

```{r gene, eval=TRUE}
# at least 10 cells should express the gene
keep_feature <- rowSums(counts(sce) > 0) > 10
sce <- sce[keep_feature,]
```

We keep `r sum(keep_feature)` genes

```{r}
if(!file.exists(here("processed", project,  "sce_QC_02.RDS"))){
  saveRDS(sce, here("processed", project,  "sce_QC_02.RDS"))
  }
```

Final dimension is `r dim(sce)`
